---
title: "MSSP 615 tnum notes"
author: "[Allen Razdow](arazdow@truenum.com)"
output: html_notebook
---

This document addresses questions and issues that came up in today's class.  

### Package documentation  
Yes, it is very minimal at this early stage in development of the package.  There are man pages for most of the functons.

For example: *> ?tnum.tagByQuery* will display that function's documentation in the help window.


### Case in REGEXP queries

Most regular expression implementations include "flags" affecting the way a pattern is used.  You can find them described in most regular expression references. Here are ones useful in tnum:

- *i*  --
With this flag the search is case-insensitive: no difference between A and a (see the example below).

- *u* --
Enables full unicode support. The flag enables correct processing of surrogate pairs. 

```{r}
# This line assumes you have tnum.authorized() already
# We pass the "i" flag to REGEXP as a second arg, also in escaped quotes:

qresult <- tnum.query("*pride* has text = REGEXP(\"darcy\", \"i\")")

# Display the value of one of the returned sentence tnums, using as.character() to strip
#  the attributes of the object, so only the text value of the sentence is shown.  
#  Note that "Darcy" was matched by "darcy" using the "i" flag.  

 as.character(qresult[[1]])

```
### Deleting tags from the database

Use the _tnum.tagByQuery_ functon _removes =_ argument.  Tags in the removes list will be deleted:

tnum.tagByQuery(query = "", adds = list(), removes = list())

### Using tnum.graphTnumList( )

This type of graph is useful because it shows subjects, properties and tags in one picture.  We tag sentences in Sense and Sensibility with two different tags to illustrate:
```{r}
# Tag sentences mentioning money
tnum.tagByQuery("*sense* has text = REGEXP(\"pounds|money\",\"i\")", adds="notes:money")

# Tag mentions of romance
tnum.tagByQuery("*sense* has text = REGEXP(\"love|romantic|romance\",\"i\")", adds="notes:love")
```

Now we can retreive sentences with either tag, and graph the result:

```{r}
taggedNums <- tnum.query("@notes#", max=200)

# Since our query returned many tnums, our graph would be very cluttered if we don't do some aggregation of nodes
# "collectors" are patterns to match segments of the subjects, with special wildcard ### on which to collect nodes.
# We use two collectors for paragraphs and sentences.  This will result in collecting tags for each chapter.  
#
# We use a regular expression to select the tags we just added, leaving out the many tags added during your projects. 
# 

tnum.graphTnumList(taggedNums,
                   collectors = list("paragraph-###","sentence-###"),
                   tagpattern = "notes:.+",
                   style = "tree"
                   )
```
Finally, we remove the tags we just used:
```{r}
tnum.tagByQuery("@notes#", removes = list("notes:love","notes:money"))
```