---
title: "Writing Data Frames as Truenumbers"
author: "Allen Razdow"
date: "5/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Comparing paradigms

Data frames in R are record-oriented data.  A sequence of rows, each a set of name-value pairs.  The names are the same for each row, comprising its columns.  Data frames are the basic data structures for analysis in R, and their form never varies regardless of the kind of the data they contain.  Only the column names and values hint at the reality the data represents, because data frames are for computation, not description.  

Truenumbers is in that sense the opposite of tabular data, because it catalogs information by description, having no other explicit structure.   

This data frame has four columns, and two rows, or *observations*:
```{r }
 df1 <- data.frame(state=c("arizona","arizona"),city=c("flagstaff","tucson"),population=c(76831,542629), meanTempF = c(71, 88))
knitr::kable(df1)
```
We can get the gist of what the frame is all about from reading the names, and figure out that the *state* and *city* columns specify the subject for which the population and average temperature are given in the other columns.  In TN, subject and property are explicit: 

 **flagstaff arizona has population = 76831**
 
 **flagstaff arizona has mean temperature = 71 F**
 
 **tucson arizona has population = 542629**
 
 **tucson arizona has mean temperature = 88 F**

There are several things to notice about this representation:

 * Instead of associating fields by putting them in the same row (*tuple* concept) TNs are associated by having the same subject phrase
 * Where the row is identified by *state* and *city* columns, the phrase is self-explanatory with no metadata: **tucson arizona**
 * Column head *meanTempF* becomes a searchable property **mean temperature** and the value has intrinsic units of measure
 * Phrases are part of the data, and not restricted to a schema, so more information is captured than in a table

## Considering subject identity

Had we built our dataframe without the *state* column, we could find ourselves with N rows in which the *city* was **Springfield.**  That is, the *city* column alone would fail to be a unique key.  However, we would still know which population and temperature numbers are together because they'll be in the same row.  In the TN case, if we had N **Springfield** population numbers, we would have N *temperature* numbers, and no way to know which ones go together.

In general, the problem is the same: insufficient identification of the subject.  Consideration of subjects and properties is intrinsic to writing TNs, and sufficient identification, whether in the subject phrase or using tags, is an important consideration.   

## Using TNUM package to write TNs from a data frame

The **tnum.ingestDataFrame** function is used to generate a set of tagged truenumbers to represent each row in the data frame.  We'll use it to generate the descriptive TNs shown earlier.  We define the list of templates, in this case, one for the population and one for the temperature. We add a tag to both indicating the state.

```{r}
templates <- list(
  c("$(city) $(state) has population = $(population)", "state:$str_to_upper(state),ingest:$dateTkn()"),
  c("$(city) $(state) has mean temperature = $(meanTempF) F", "state:$(state)")
)
```

```{r , include=FALSE}
# Load TNUM and authenticate to access the server
library(tnum)
tnum.loadLibs()
tnum.authorize(ip="metrics.truenum.com:8080",creds = "demo@truenum.com:pinecone")

```

Now we call the ingest function, using the *outfile* parameter to write the generated TNs to a file suitable for uploading using the TN web dashboard, and display the file. That way, we're not actually posting anything.

```{r}
tnum.ingestDataFrame(df1, templates, "testIngest1.txt")

read.delim("testIngest1.txt", header = FALSE)

```
Now we ingest them again, posting to the numberspace:
```{r}
tnum.ingestDataFrame(df1, templates)
```
We query for them by tag, and display the resulting JSON as a dataframe
```{r}
results <- tnum.query("state:arizona")
knitr::kable(results)
```
