---
title: "Getting Started wth **tnum** for R"
author: "[True Engineering Technology, LLC](http://truenum.com) Lincoln, MA" 
output: html_notebook
---

***

### 

***tnum* is a data management companion for R.**  In using R, we draw from a variety of data sources, reading them into data frames, vectors or lists for exploration and analytic processing.  This in turn produces more data, from summary statistics and models to graphs and tables that enrich the original data, or derive new information from it. Doing this results in many different data structures, labeled with informal row and column names, and many files persisting source data and results on the user's hard-drive. Working and collaborating with these requires keeping track of them, their meaning, structure and file versions. 

**Truenumbers**, is a central cloud repository for creating, tagging and querying persistent data as a collection of facts using a structured pseudo-natural language to describe them. We call these facts *truenumbers*, *tnums* or *TNs*.  Using *tnum* provides a convenient and scalable way to manage data and results centrally, as a growing body of intelligible knowledge, without resorting to relational databases (which are usually far from intelligible).  


### Getting up and running

After installing package *tnum*, it's necessary to authenticate with the *tnum* server:
```{r}
library(tnum)
tnum.authorize(ip="metrics.truenum.com:8080",creds = "demo@truenum.com:pinecone")
```
*authorize* takes two arguments: **ip** representing the URL or IP address of the *tnum* server endpoint.  **creds** is a string containiny your username and password, separated b a colon. The function displays a list of available numberspaces on the server. A *numberspace* is a separate repository for truenumber data.  The authorize() call sets the current numberspace to a numberspace defined by your administrator.    

### Truenumber facts and phrases

A truenumber fact (a *tnum*) is equivalent to a sentence of the form **&lt;subject&gt; has &lt;property&gt; = &lt;value&gt;**, or **&lt;property&gt; of &lt;subject&gt; is &lt;value&gt;** for example:

<br>
<center>
**number of fulltime undergraduate students of colorado state university is 15646**
</center>
<br>
and
<br>
<center>
**montclair state university has graduation rate = 58%**
</center>
<br>
The subject  **fulltime undergraduate students of montclair state university ** is equivalent to a path **university:state:montclair/students:undergraduate:fulltime** where the separator **:** denotes an *adjective* relationship, and **/** denotes a possesive.

Let's make truenum JSON objects for them:

```{r}
tn1 <- tnum.postStatement(
  stmt = "number of fulltime undergraduate students of colorado state university is 15646",
  notes = "Our first truenumber"
)

tn2 <- tnum.postStatement( stmt = "montclair state university has graduation rate = 58%", notes = "Our second truenumber")
```

```{r}
df <- tnum.jsonArray(list(tn1,tn2))

kable(df[,1:7])
```
```{r}
kable(df[,8:length(df)])

```

### Visualizing description trees

